The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-07-18 paper.tex
Finished Irish SonAngle section

Epenthesis does not break up onset sonorant-initial clusters.  This is likely due to the fact that
with a small number of lexical exceptions, Irish words are stress-initial \citep[26]{o.siadhail.1989}.\footnote{With the exception of Munster Irish, where stress is attracted to syllables containing
a long vowel or diphthong (Green 1996), but where onset clusters still are not broken up by
epenthesis.  In this case, one would probably have to posit a positional faithfulness constraint
protecting the word-initial cluster.}  
Breaking up the onset cluster would result in an unstressable epenthetic vowel occupying the 
preferred location of stress in Irish.

@@

>>>>

Epenthesis does not break up onset sonorant-initial clusters.  This is likely due to the fact that
with a small number of lexical exceptions, Irish words are stress-initial \citep[26]{o.siadhail.1989}.\footnote{With the exception of Munster Irish, where stress is attracted to syllables containing
a long vowel or diphthong (Green 1996), %TODO
but where onset clusters still are not broken up by
epenthesis.  In this case, one would probably have to posit a positional faithfulness constraint
protecting the word-initial cluster.}  
Breaking up the onset cluster would result in an unstressable epenthetic vowel occupying the 
preferred location of stress in Irish.

@@

In codas and across syllable boundaries, epenthesis does not apply to homorganic clusters.  
I suggest that this is due to the action of the Obligatory Contour Principle (OCP).  
I adopt Walter 2007's %TODO
formulation of the OCP, which bans successive identical oral gestures.  A cluster such as [nt] would consist of a single oral constriction gesture at the alveolar ridge, and would not constitute a violation of the OCP. Splitting it by epenthesis to form [n\textipa{@}t] would require a repeated identical constriction gesture, and therefore be a violation of the OCP.

\ex. OCP $\gg$ {\sc *Son-C}

\vspace{-2em}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{bord}/} & OCP & {\sc *Son-C} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{bord} & & \cellcolor{lightgray}$\ast$ \\
\hline b. & & \textipa{bor@d} & $\ast$! & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

Lastly, we come to the sonority-driven aspect of the analysis.  
Epenthesis is allowed when C1 is ``followed by a voiced stop, fricative, nasal or liquid, but not when it is followed by a voiceless stop.'' \citep{carnie.1994}

This requires a revision to the standard sonority scale, splitting up voiceless and voiced stops:

\ex.  \begin{center}
   \begin{tabular}{cccccccc}
    Voiceless stop & Voiced stop  & Fricative    & Nasal        & Liquid & Glide        & Vowel \\
      T            &     D       &          F    &    N &    R   &   G          & V     \\
      1            &     1.5     &          2   &    3  &    4   &   5          & 6     \\
     \end{tabular}
 \end{center}

Now let us calculate the sonority angles for the relevant set of sonorant-initial clusters in Irish.

\ex. {\sc Sonority Angles} of sonorant-initial clusters using revised sonority scale

  \begin{tikzpicture}[shorten >=1pt,->,scale=0.5]
     \tikzstyle{line} = [draw]%, -latex']

        \node (NT) at (2.48619449019234 * 15, 0) {NT}; % 2.35619449019234 - repositioning for visual clarity
        \node (ND) at (2.28183949564558 * 15, 0) {ND}; % 2.23183949564558 
        \node (NF) at (2.03444393579570 * 15, 0) {NF};
        \node (NN) at (1.24904577239825 * 15, 0) {NN};
        \node (NR) at (0.60 * 15, 0) {NR}; % 0.4636476090008

        \node (RT) at (2.48619449019234 * 15, 0.7) {RT};% 2.35619449019234 - repositioning for visual clarity
        \node (RD) at (2.37743866747662 * 15, 0) {RD}; % 2.29743866747662 
        \node (RF) at (2.19429743558818 * 15, 0) {RF}; % 2.21429743558818 
        \node (RN) at (1.89254688119154 * 15, 0) {RN};
        \node (RR) at (1.10714871779409 * 15, 0) {RR};

    \node (axisstart) at (0.5 * 15,-1) {};
    \node (axisend)   at (2.6 * 15,-1) {};
    \draw (axisstart) -- (axisend);
   \node (xaxislabel) at (2.6 * 15,-2.5) {\textsc{Sonority Angle}};

  \node (dividinglinestart) at (2.43 * 15,2) {};
  \node (dividinglineend)   at (2.43 * 15,-2) {};
  \path [line,dashed] (dividinglinestart) -- (dividinglineend) -- cycle;

  \node (2pointlabel) at (2.43 * 15,-1.5) {2.3};

    \end{tikzpicture} %TODO : draw arrow and mark 1.0, 1.5, 2.0, 2.5

The sonorant-voiceless stop clusters can be neatly separated from the rest of the clusters in terms of {\sc Sonority Angle}.
The correct constraint ranking is therefore the following:

\ex. {\sc Ident(SonAngle)$<$2.3} $\gg$ {\sc *Son-C} $\gg$ {\sc Ident(SonAngle)$<$2.4}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{kork}/} & {\sc Ident(SonAngle)$<$2.3} & {\sc *Son-C} & {\sc Ident(SonAngle)$<$2.4} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{kork} & & \cellcolor{lightgray}$\ast$ & \cellcolor{lightgray} \\
\hline b. & & \textipa{kor@k} & 2.35 $\ast$! & \cellcolor{lightgray} & \cellcolor{lightgray}2.35 $\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{alb@}/} & {\sc Ident(SonAngle)$<$2.3} & {\sc *Son-C} & {\sc Ident(SonAngle)$<$2.4} \\[0.5ex]
\hline \hline c. & \ding{43} & \textipa{al@b@} & 2.29 & & \cellcolor{lightgray} 2.29 \\
\hline d. & & \textipa{alb@} & & $\ast$! & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

A note on the nasal-voiceless stop clusters is in order here: such clusters tend to be homorganic on the surface.  
It is therefore unclear whether such clusters failed to be broken up by epenthesis because of OCP, 
or because of the large sonority contour faithfulness violation.  
I argue that in general, it is for the second reason.

Suppose nasal-voiceless stop clusters were banned from epenthesising because of the OCP.
Then we would expect to see surface
sequences of [NVP] where N and P have different places of articulation and V is an epenthetic vowel, 
just as we see [NVB] sequences:

\ex. \a. [\textipa{b\textsuperscript{j}in\textsuperscript{j}ib}] \\
         {\it binb} `venom'
     \b. [\textipa{ban@b@}] \\
         {\it Banba} `Banba (a name for Ireland)'
     \z.
     \citep[2c]{carnie.1994}

To my knowledge, no [NVP] sequences with the properties stated above exist.

Further, albeit marginal, evidence comes from the existence of words with heterorganic [NP] sequences,
which are not broken up by epenthesis.

\ex. \a. [\textipa{s\textsuperscript{j}e:mt}] \\
         {\it s\'eimt} `the act of playing music', dialectal variant of {\it seinm})
     \b. [\textipa{k2r\textsuperscript{j}@m\textsuperscript{j}k\textsuperscript{j}}] \\
         {\it cuirimc} `I observe, pay attention to' (Connaught dialect) \\
     (Words found in \citet{dinneen.2007}) %TODO

I conclude that in general, epenthesis does not occur in /NT/ sequences due to the large sonority angle between NT and NVT, and not to the non-existence of underlying heterorganic /NP/ clusters in the lexicon.
